---
title: "Air quality in Madrid (Spain) - From 2001 to 2018"
author: "Wagner Rosa"
output: 
  pdf_document: default
  html_document: default
---
# Introduction
Air pollution in the big metropoles around the world is one of the big concerns nowadays, since we can observe the negative effects of the pollutants in health and climate. Therefore, a control of the particles and gases with high potential to be harmful to our environment is needed. For that matter, Madrid is one of Europe's big capitals and has a series of air quality control stations distributed around the city, which can provide us a nice piece of information (and lots of data) about the pollutants presented at the Spanish capital. As a spanish myself and have lived in Madrid for many years, this topic is of very importance personally. 

Recently, Madrid has launch the campaing "Madrid Central" in 2018, [Wikipedia](https://en.wikipedia.org/wiki/Madrid_Central), which determine what type of vehicles are permitted to traffic in the capital's main area, depending on the year of the vehicle, motorization and so on. For instance, Madrid expect to reduce the pollution on Central area significantly. Hopefully, in this Capstone project, we will be able to predict if the ongoing trend is already a decline on pollutants or, worst case scenario, an increasing of pollutant in the spanish capital.

All the data sets were collected from [Kaggle](https://www.kaggle.com/decide-soluciones/air-quality-madrid) 

# Loading the necessary libraries
```{r}
#Loading the necessary libraries
if(!require(tidyverse)) install.packages("tidyverse", repos = "http://cran.us.r-project.org")
if(!require(caret)) install.packages("caret", repos = "http://cran.us.r-project.org")
if(!require(data.table)) install.packages("data.table", repos = "http://cran.us.r-project.org")
if(!require(lubridate)) install.packages("lubridate", repos = "http://cran.us.r-project.org")
#if(!require(anova)) install.packages("anova", repos = "http://cran.us.r-project.org")
```
Now let's load the CSV files
```{r}
#Setting the right directory path and importing csv files
setwd("~/Library/Mobile Documents/com~apple~CloudDocs/Data Science/RProjetcs/Capstone Project/air-quality-madrid")
stations.code <- read.csv("stations.csv", sep = ",")
```

# Exploratory data analysis (EDA)

```{r}
glimpse(stations.code)
```

```{r}

#Assigning the list of csv files to a variable
files <- list.files(
  path = "./csvs_per_year",
  pattern = "*.csv", 
  full.names = T)

#Looping through the files and joining all the data sets to a single data frame
madrid.airquality <- sapply(files, read_csv, simplify=FALSE) %>% bind_rows

```
Looking into the file data frame format

```{r}
str(madrid.airquality, give.attr=F)
```
Getting the names from the columns for each data frame.

```{r}
colnames(madrid.airquality) 
```
The names are related to the following measurements:
- SO_2: sulphur dioxide level measured in μg/m³. High levels of sulphur dioxide can produce irritation in the skin and membranes, and worsen asthma or heart diseases in sensitive groups.

- CO: carbon monoxide level measured in mg/m³. Carbon monoxide poisoning involves headaches, dizziness and confusion in short exposures and can result in loss of consciousness, arrhythmias, seizures or even death in the long term.

- NO: nitric oxide level measured in μg/m³. This is a highly corrosive gas generated among others by motor vehicles and fuel burning processes.

- NO_2: nitrogen dioxide level measured in μg/m³. Long-term exposure is a cause of chronic lung diseases, and are harmful for the vegetation.

PM25: particles smaller than 2.5 μm level measured in μg/m³. The size of these particles allow them to penetrate into the gas exchange regions of the lungs (alveolus) and even enter the arteries. Long-term exposure is proven to be related to low birth weight and high blood pressure in newborn babies.

- PM10: particles smaller than 10 μm. Even though the cannot penetrate the alveolus, they can still penetrate through the lungs and affect other organs. Long term exposure can result in lung cancer and cardiovascular complications.

- NOx: nitrous oxides level measured in μg/m³. Affect the human respiratory system worsening asthma or other diseases, and are responsible of the yellowish-brown color of photochemical smog.

- O_3: ozone level measured in μg/m³. High levels can produce asthma, bronchytis or other chronic pulmonary diseases in sensitive groups or outdoor workers.

- TOL: toluene (methylbenzene) level measured in μg/m³. Long-term exposure to this substance (present in tobacco smkoke as well) can result in kidney complications or permanent brain damage.

- BEN: benzene level measured in μg/m³. Benzene is a eye and skin irritant, and long exposures may result in several types of cancer, leukaemia and anaemias. Benzene is considered a group 1 carcinogenic to humans by the IARC.

- EBE: ethylbenzene level measured in μg/m³. Long term exposure can cause hearing or kidney problems and the IARC has concluded that long-term exposure can produce cancer.

- MXY: m-xylene level measured in μg/m³. Xylenes can affect not only air but also water and soil, and a long exposure to high levels of xylenes can result in diseases affecting the liver, kidney and nervous system (especially memory and affected stimulus reaction).

- PXY: p-xylene level measured in μg/m³. See MXY for xylene exposure effects on health.

- OXY: o-xylene level measured in μg/m³. See MXY for xylene exposure effects on health.

- TCH: total hydrocarbons level measured in mg/m³. This group of substances can be responsible of different blood, immune system, liver, spleen, kidneys or lung diseases.

- CH4: methane level measured in mg/m³. This gas is an asphyxiant, which displaces the oxygen animals need to breath. Displaced oxygen can result in dizzinnes, weakness, nausea and loss of coordination.

- NMHC: non-methane hydrocarbons (volatile organic compounds) level measured in mg/m³. Long exposure to some of these substances can result in damage to the liver, kidney, and central nervous system. Some of them are suspected to cause cancer in humans.

As pointed in an article from [Iberdrola](https://www.iberdrola.com/medio-ambiente/contaminacion-calidad-aire) - in spanish - the control standard is usally done measuring the levels of NO2, O3, CO, SO2. We are going to also look into PM10 and PM25, which are related to particles smaller than 10 adn 2.5 μm, respectively. This is because the long exposure to these particles can can result in lung cancer and cardiovascular complications (PM10); low birth weight and high blood pressure in newborn babies (PM25). 

Other reason to choose these level is because they are presented in all the data frames.

Summary of the pollutant measurments through the years

```{r}
summary(madrid.airquality) 
```
Getting the date right

```{r}
#Ordering by date
madrid.airquality <- madrid.airquality[with(madrid.airquality, order(date)),]
madrid.airquality$full.date.time <- as.POSIXct(madrid.airquality$date,format = "%Y-%m-%d %H:%M:%S", tz='CET')
madrid.airquality$date <- format(madrid.airquality$full.date.time, "%Y-%m-%d")
madrid.airquality$time <- format(madrid.airquality$full.date.time, "%T")
madrid.airquality$day <- format(madrid.airquality$full.date.time, "%d")
madrid.airquality$month <- format(madrid.airquality$full.date.time, "%m")
madrid.airquality$year <- format(madrid.airquality$full.date.time, "%Y")
```

Getting the total number of NA's in the data frame

```{r}
#Total number of NA's
sum(is.na(madrid.airquality))
```
Now, let's filter the data to plot the amount of NA's for our selected indicators. For this, we will plot a graph to see the percentage of NA's in the data frame subset

```{r}
indicators <- c("NO_2", "O_3", "SO_2", "CO", "PM10", "PM25")

#filtering only the subset of pollutants we want to analyze
madrid.airquality.indicators <- madrid.airquality[, indicators]

na.indicators <- data.frame(percent=round(
  colSums(
    is.na(madrid.airquality.indicators))/nrow(madrid.airquality.indicators)*100)
  )

na.indicators
```
Let's plot it into a bar graph to visualize the variation in NA's percentage for each pollutant measured in this historical series

```{r}
na.indicators$pollutant <- rownames(na.indicators)
na.indicators$pollutant<-factor(na.indicators$pollutant, 
                                as.character(na.indicators$pollutant))

na.indicators %>% 
  mutate(pollutant = fct_reorder(pollutant, percent)) %>%
  ggplot(aes(pollutant, percent,fill = pollutant)) +
  geom_bar(stat = "identity") + 
  scale_fill_brewer() +
  geom_text(aes(label=paste0(percent,"%"), y=percent+0.7), 
                size = 3,
                hjust = -0.01
                ) +
  labs(x = "Pollutants", 
       y = "Percentage", 
       title = "Percentage of NAs in Four Pollutants") +
  coord_flip()
```

PM25 has a considerable large number of NA's with 79% rate. The other pollutants have values between 20 and 30 percent, with only exeception of NO2 (1%)

Now, let's take a look on how these pollutants/indicators vary over time.
```{r}
#Daily variation
madrid.daily <- madrid.airquality %>%
  group_by(date) %>%
  summarise(NO_2.mean = mean(NO_2, na.rm = T),
            O_3.mean = mean(O_3, na.rm = T),
            PM10.mean = mean(PM10, na.rm = T),
            SO_2.mean = mean(SO_2, na.rm = T),
            CO.mean = mean(CO, na.rm = T),
            PM25.mean = mean(PM25, na.rm = T))

#Getting the total pollutant
madrid.daily <- mutate(madrid.daily, 
                       tot_poll = rowSums(madrid.daily[,-1], 
                       na.rm = T))

#Plottind the data frame
madrid.daily %>% 
  ggplot(aes(x = as.Date(date), 
             y = tot_poll)) + 
    geom_line(color = 'blue') +
    theme(legend.position = "none",
          panel.background = element_rect(fill = "gray94", colour = "white",
                                      size = 0.5, linetype = "solid"),
          legend.background = element_rect(fill = "gray94"),
          plot.background = element_rect(fill = "gray94"),
          panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour =                                            "white"), 
          panel.grid.minor = element_line(size = 0.25, linetype = 'solid', colour =                                           "white"),  
          plot.title = element_text(hjust = 0, face = 'bold',color = 'black')) +
    labs(x = '', 
         y = 'Total Pollutants (μg/m³)',
         itle='Air Quality in Madrid', 
         subtitle='Indicators: NO2, O3, PM10, SO2, CO, PM25') +
    theme(axis.text.y =element_text(angle = 0,
                                    size = 8),
          axis.text.x=element_text(angle = 90, 
                                   size = 8)) +
    scale_x_date(breaks = seq(as.Date("2001-01-01"), 
                              as.Date("2018-07-01"), 
                              by="12 months"), 
                 date_labels = "%Y")
```

To look how the average total pollution varies through the months, we're going to use the same approach as the one applied for the daily variation:
```{r}
#Monthly variation
madrid.monthly <- madrid.airquality %>%
  group_by(year, month) %>%
  summarise(NO_2.mean = mean(NO_2, na.rm = T),
            O_3.mean = mean(O_3, na.rm = T),
            PM10.mean = mean(PM10, na.rm = T),
            SO_2.mean = mean(SO_2, na.rm = T),
            CO.mean = mean(CO, na.rm = T),
            PM25.mean = mean(PM25, na.rm = T)) %>%
  mutate(time = paste(year, "-", month, "- 01", sep = ""))

#Getting the total pollutant
madrid.monthly$tot_poll_mon <- rowSums(madrid.monthly[,3:8], na.rm = T) # this selects only the columns with the averages. In order to check it out, uncomment the next code line:  
#str(madrid.monthly)
madrid.monthly$time_mon = as.Date(madrid.monthly$time, format = "%Y-%m-%d")

#Plottind the data frame
madrid.monthly %>% 
  ggplot(aes(x = time_mon, 
             y = tot_poll_mon)) + 
    geom_line(color = 'black') +
    theme(legend.position = "none",
          panel.background = element_rect(fill = "gray94", colour = "white",
                                      size = 0.5, linetype = "solid"),
          legend.background = element_rect(fill = "gray94"),
          plot.background = element_rect(fill = "gray94"),
          panel.grid.major = element_line(size = 0.5, linetype = 'solid', colour =                                            "white"), 
          panel.grid.minor = element_line(size = 0.25, linetype = 'solid', colour =                                           "white"),  
          plot.title = element_text(hjust = 0, face = 'bold',color = 'black')) +
    labs(x = '', 
         y = 'Total Pollutants (μg/m³)',
         itle='Air Quality in Madrid', 
         subtitle='Indicators: NO2, O3, PM10, SO2, CO, PM25') +
    theme(axis.text.y =element_text(angle = 0,
                                    size = 8),
          axis.text.x=element_text(angle = 90, 
                                   size = 8)) +
    scale_x_date(breaks = seq(as.Date("2001-01-01"), 
                              as.Date("2018-07-01"), 
                              by="6 months"), 
                 date_labels = "%b-%y") +
    scale_y_continuous(breaks = seq(0,250,50), lim = c(0,250))
```
This time series looks like a smooth out curve compared with the daily averages, as expected.

I need:
clean na's
set date in right format
look at whole pollutants through the years
see how selected pollutants evolved through the years
see how (during  a year) the concentration of pollutants vary (seasonal effects) - find patterns
look how the pollutants vary from station to station
split the data set into train/test/validation
check arima, anova and random forest approximations